Drought, vegetation productivity, and land cover change in Chile

### 2024-03-12 article/paper_drought_Chile2023.tex
update version1 article

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, boxrule=0pt, frame hidden, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

\hypertarget{earth-observation-data}{%
\subsubsection{Earth observation data}\label{earth-observation-data}}

For water supply and demand variables, we used ERA5L
\citep{MunozSabater2021}, a reanalys dataset that provides the evolution
of land variables since 1950. It has a spatial resolution of 0.1°,
hourly frequency, and global coverage. We selected the variables for
total precipitation, 2 meter temperature maximum and minimum, and
volumetric soil water layers between 0 and 100cm of depth (layer 1 to
layer 3). The data was downloaded using the Copernicus Climate Data
Store (CDS) Application Program Interface (API) implemented in
\texttt{\{ecmfwr\}} \citep{Hufkens2019}.

To derive a proxy for vegetation productivity, we used the product
MOD13A3 collection 6.1 from MODIS \citep{Didan2015}. It provides
vegetation indices (NDVI and EVI) at 1km of spatial resolution and
monthly frequency. The MOD13A3.061 and MCD12Q1.061 were retrieved from
the online Data Pool, courtesy of the NASA EOSDIS Land Processes
Distributed Active Archive Center (LP DAAC), USGS Earth Resources
Observation and Science (EROS) Center, Sioux Falls, South Dakota,
https://lpdaa.usgs.gov/tools/data-pool/.

@@

\hypertarget{weather-stations}{%
\subsubsection{Weather stations}\label{weather-stations}}

We compared the ERA5L variables for monthly mean temperature, total
precipitation, and volumetric soil water content against values
retrieved by weather stations. For temperature and precipitation, we
used the weather network from the Ministry of Agriculture of Chile
(www.agromet.com) between 2015 and 2023. We used 277 stations located
throughout Chile. For soil moisture, we select a private soil network
that is owned by the agricultural enterprise Garces Fruit, which has 99
stations in Central Chile, located in cherry fruit crops. The sensors
are installed at 30, 60, and 90m and are the model Teros 12 from
MeterGroup. To avoid the effect of irrigation on soil moisture, which
ERA5L hardly captures, we used daily data for the year 2022 and the
months outside the growing season, May to September.

\hypertarget{validation-of-era5l-variables}{%
\subsubsection{Validation of ERA5L
variables}\label{validation-of-era5l-variables}}

To account for the performance of the ERA5L climatic variables regarding
the values measured by the weather stations. We selected the following
metrics:

\[MAE = \frac{1}{n}\sum |{E-S}|\] \[Bias = \frac{\sum E}{\sum S}\]
\[ubRMSE =\sqrt{\frac{\sum{ \left[ (E_i-\overline{E})-(S_i-\overline{S}) \right ] ^2}}{n}}\]

\[CC = \frac{\sum (S_i-\bar{S})(E_i-\bar{E})}{\sqrt{(Si-\bar{S})^2(E_i-\bar{E})^2}}\]

\(MAE\): mean absolute error \(bias\): bias \(ubRMSE\): unbiassed root
mean squared error \(CC\): coefficient of correlation \(S\): value of
the variable measure by the weather station \(E\): value of the variable
measure by ERA5L

\hypertarget{drought-indices}{%
\subsection{Drought Indices}\label{drought-indices}}

@@

For the indices EDDI and SPEI that use water demand, first we have to
calculate the AED. For this, we used the method of Hargreaves
\citep{Hargreaves1994, Hargreaves1985}:

@@

We chose the method of Hargreaves to estimate AED because of its
simplicity, which only requires temperatures and extrarrestrial
radiation. Also, it has been recommended over other methods when the use
of several climatic variables is limited \citep{Vicente-Serrano2014}.

@@

We derived the drought indices of water supply and demand, soil moisture
from the ERA5L dataset, and vegetation from the MODIS product, all at
monthly frequency.

To evaluate water demand, we chose the \(EDDI\)
\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). We used
the SPI \citep{Mckee1993}, a drought monitoring index that the World
Meteorological Organization (WMO) recommends. We calculated the SPEI,
which used a balance between \(P\) and \(AED\), in this case, an
auxiliary variable \(D = P-AED\) is used. In this study, we used the
\(SSI\) (standardized soil moisture index at 1 m)
\citep{Hao2013, AghaKouchak2014}, which uses soil moisture at 1m depth.
Finally, for the proxy of productivity, \(zcNDVI\), we used the NDVI.
Before using the NDVI, it was smoothed using a locally weighted
polynomial regression, following the procedure described in
\citet{Zambrano2018} and \citet{Zambrano2016}.

All the indices are multi-scalar and were calculated for time scales of
1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which was calculated
for 6 months. The goal is to be able to evaluate short- and long-term
droughts in water demand and supply and soil moisture. This is
particularly important for central Chile because it has suffered from a
prolonged decrease in precipitation for more than 12 years
\citep{Garreaud2020, Boisier2018, Garreaud2017}.

To calculate the drought indices, first we must calculate the
accumulation of the variable. In this case, for generalization purposes,
we will use \(V\), referring to \(P\), \(AED\), \(D\), \(NDVI\), and
\(SM\) (Table \ref{tab-desEOD}). We cumulated each \(V\) over the time
series of \(n\) values, and for the time scales \(s\):

@@

It corresponds to a moving window (convolution) that sums the variable
for \(s\) starting for the last month \(n\) until the month, which could
sum for \(s\) months (n-s+1). Once the variable is cumulated over time
for the scale \(s\), we used a nonparametric approach following
\citet{Hobbins2016} to derive the drought indices. Thus, the empirically
derived probabilities are obtained through an inverse normal
approximation \citep{Abramowitz1968}. Then, we used the empirical Tukey
plotting position \citep{Wilks2011} over \(A_i\) to derive the
\(P(A_i)\) probabilities across a period of interest:

@@

\hypertarget{sec-methods_lulc}{%
\subsection{LULC change for 2001-2022 and its relation with water supply
and demand, and soil moisture}\label{sec-methods_lulc}}

\hypertarget{land-cover-macroclasess-and-validation}{%
\subsubsection{Land cover macroclasess and
validation}\label{land-cover-macroclasess-and-validation}}

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
The IGBP defines 17 classes; from these, we regrouped into ten
macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9
to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands,
13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
Thus, we have a land cover raster time series with the ten classes for
2001 and 2023.

To validate the land cover obtained, we compare the macroclasses with
the ones of a more detailed land cover map made by \citet{Zhao2016} for
Chile with samples acquired in the years 2013--2014 (LCChile). The later
has a spatial resolution of 30 m and three levels of defined classes;
from those, we used level 1, which fits with the macroclasses land
cover. We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from land
cover macrolcasses to validate with LCChile.

We follow the next procedure:

\begin{enumerate}
\def\labelenumi{\roman{enumi})}
\tightlist
\item
  resampled LCChile to the spatial resolution (500m) of the land cover
  macroclasses using the nearest neighbor method,
\item
  took a random sample of 1000 points within continental Chile and
  extracted the classes that fell within each point for LCChile,
  IGBP2013, and IGBP2014; we considered the point extracted from LCChile
  as the truth and the values from the other two years as predictions.
\item
  calculate a confusion matrix with the classes extracted from the 1000
  points for LCChile, IGBP2013, and IGBP2014. Calculate the performance
  metrics of accuracy and F1.
\end{enumerate}

\[Accuracy = \frac{TP+TN}{TP+TN+FP+FN}=\frac{correct\,\, classifications}{all\,\, classifications}\]
\[F1=\frac{2\cdot TP}{2\cdot TP + FP +FN}\] where \(TP\) and \(FN\)
refer to true positive and false negative, correctly classified classes;
\(TN\) and \(FP\) to true negative and false positive, wrongly
classified classes.

\hypertarget{land-cover-persistence-mask-2001-2022}{%
\subsubsection{Land cover persistence mask
2001-2022}\label{land-cover-persistence-mask-2001-2022}}

Climate, vegetation development, seasonality, and changes in vegetation
type all have an impact on the time series of NDVI. In this study, we
want to examine the variation in vegetation productivity across various
land cover types and how water demand, water supply, and soil moisture
affect it. In order to avoid changes due to a change in the land cover
type that will wrongly impact NDVI, we will develop a persistence mask
for land cover for 2001--2023. Thereby, we reduce an important source of
variation on a regional scale.

Therefore, we generated a raster mask for IGBP MODIS per pixel using
macroclasses that remain constant for at least 80\% of the years
(2001--2022). This enables us to identify regions where the land cover
of the macroclasses has remained constant.

\hypertarget{land-cover-trend-and-drought-indices}{%
\subsubsection{Land cover trend and drought
indices}\label{land-cover-trend-and-drought-indices}}

We calculated the surface occupied per land cover class into the five
macrozones (``Norte Grande'' to ``Austral'') per year for 2001--2023.
After that, we calculated the trend's change in surface; we used the
Sen' slope \citep{Sen1968} based on Mann-Kendall \citep{Kendall1975}.
This way, we obtain a matrix of trends of 5 x 5 (macrozones x land
cover). The aim is to later explore if the trend in land cover classes
is associated with a trend in the drought indices. For this, we will use
the techniques of regresion and regularization of Lasso
\citep{Tibshirani2010} and Ridge \citep{Hoerl1970}. Also, we will test
random forests for this purpose \citep{Ho1995}. We will choose the trend
of land cover surface per macroclass and macrozone as the response
variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI
for time scales 1, 3, 6, 12, 24, and 36 months) as the predictor
variables. With this analysis, we expect to gather insights regarding
whether there is a pattern of climatic influence along Chile or if what
is happening in Central Chile has to do with more localized climatic
conditions.

\hypertarget{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}{%
\subsection{Trend of drought indices for water demand and supply, soil
moisture, and vegetation
productivity}\label{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}}

\hypertarget{mann-kendall-and-sens-slope}{%
\subsubsection{Mann-Kendall and Sen's
slope}\label{mann-kendall-and-sens-slope}}

To estimate if there are significant positive or negative trends for the
drought indices, we used the non-parametric test of Mann-Kendall
\citep{Kendall1975}. To determine the magnitude of the trend, we used
Sen's slope \citep{Sen1968}. Some of the advantages of applying this
methodology are that the Sen's slope is not affected by outliers as
regular regression does, and it is a non-paramteric method that is not
affected by the distribution of the data. We applied both to the six
time scales from 1981 to 2023 (monthly frequency) and the indices SPI,
EDDI, SPEI, and SSI. In the case of zcNDVI (six months) was for 2000 to
2023. Thus, we have 31 trends. Also, we extracted the trend aggregated
by macrozone and land cover class, obtaining a table of 31x5x5 (drought
indices trends x macrozone x land cover class). We will use this data in
Section~\ref{sec-methods_lulc} to analyze if there is a strong
relationship between the trends of drought indices and land cover
surface within continental Chile.

\hypertarget{trend-in-vegetation-productivity-without-land-cover-change}{%
\subsubsection{Trend in vegetation productivity without land cover
change}\label{trend-in-vegetation-productivity-without-land-cover-change}}

\citet{Vicente-Serrano2021} made a global analysis of the drought's
severity trend using SPI, SPEI, and the Standardized Evapotranspiration
Deficit Index (SEDI; \citet{Vicente-Serrano2018}) to evaluate AED. They
indicate that the increase in hydrological drought has been due to
anthropogenic effects rather than climate change. This is because the
global increase in AED did not explain the change in the spatial pattern
of the hydrological drought. Also, they state that \emph{``the increase
in hydrological droughts has been primarily observed in regions with
high water demand and land cover change''}. We will contrast this
hypothesis with what is occurring in Chile. To achieve this, we will use
the land cover class type that remains more than 80\% of types for
2001--2022 to evaluate the trend on zcNDVI and use this as a mask where
there are low changes.

\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-land-cover-types}{%
\subsection{Impact for water supply and demand, and soil moisture in
vegetation productivity within land cover
types}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-land-cover-types}}

We analyze the drought indices of water demand and supply and soil
moisture against vegetation to address: i) if short- or long-term time
scales are most important in impacting vegetation through Chile; and ii)
the strength of the correlation for the variable and the time scale.
Then, we will summarize for each land cover class and macrozone. Thus,
we will be able to advance in understanding how climate is affecting
vegetation, considering the impact on the five macroclasses of
vegetation: forest, cropland, grassland, savanna, and shrubland.

An analysis is conducted on the linear correlation between the indices
SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36
months, and zcNDVI. The objective is to determine the impact of soil
moisture and water demand and supply on vegetation productivity. We
implemented a methodology similar to that of \citet{Meroni2016} when
comparing the SPI for meteorological drought to the cumulative FAPAR
(Fraction of Absorbed Photosynthetically Active Radiation), which served
as an indicator for vegetation productivity. A pixel-to-pixel linear
correlation analysis was performed for each index. To begin, the Pearson
coefficient of correlation is computed for each of the six time scales.
A significant time scale is identified as the one that attains the
highest correlation (p \textless{} 0.05). Subsequently, the Pearson
correlation coefficient corresponding to the time scales at which the
value peaked was extracted. As a result, for each index, we generated
two raster maps: one containing the time scales and the other the
correlation value.

@@

\hypertarget{data-1}{%
\subsection{Data}\label{data-1}}

\hypertarget{validation-of-era5l-variables-1}{%
\subsubsection{Validation of ERA5L
variables}\label{validation-of-era5l-variables-1}}

The average metrics of performance of ERA5L over the 266 weather
stations were in the case of monthly temperature: \(ubRMSE=1.06\,°C\),
\(MAE=1.131\,°C\), and \(CC=0.963\), showing a good agreement, low
error, and low overestimation. For cumulative monthly precipitation,
\(MAE=28.1\,mm\), \(bias=1.93\), and \(CC=0.845\), showing a high
correlation and a 93\% bias and being overestimated by ERA5L. In the
case of the 97 soil moisture stations, we averaged for the three depths
(30, 60, and 90m) and then compared it with volumetric water content at
1m derived from ERA5L. For this case, we made a daily comparison, having
a \(CC=0.71\), \(RMSE=0.174\,m^3m^{-3}\), \(MAE=0.167\,m^3m^{-3}\), and
\(bias=1.74\). The ERA5 soil moisture overestimate is 74\%, but it has a
kind of good correlation. For more detailed information, consult the
supplementary material (SSX).

\hypertarget{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}{%
\subsection{LULC change for 2001-2022 and its relation with water supply
and demand, and soil
moisture}\label{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}}

\hypertarget{land-cover-macroclasess-and-validation-1}{%
\subsubsection{Land cover macroclasess and
validation}\label{land-cover-macroclasess-and-validation-1}}

For vegetation, we obtained and use hereafter five macroclasses of land
cover from IGBP MODIS: forest, shrubland, savanna, grassland, and
croplands. Figure~\ref{fig-studyArea} c shows the spatial distribution
of the macroclasses through Chile for the year 2022. The validation of
IGBP2013 and IGBP2014 with LCChile reached near the same metrics of
performance, having an accuracy of \textasciitilde0.82 and a F1 score of
\textasciitilde0.66 (see SS1).

\hypertarget{sec-persistence_mask}{%
\subsubsection{Land cover persistence mask
2001-2022}\label{sec-persistence_mask}}

Figure~\ref{fig-studyArea} d, shows the macroclasses of land cover
persistance (80\%) during 2021-2022, respectively
(\ref{tab-landcoverSurf}). Within continental Chile, forest is the
vegetation type with the highest surface area at 135,00 \(km^2\),
followed by grassland (73,176 \(km^2\)), savanna (54,410 \(km^2\)),
shrubland (24,959 \(km^2\)), and cropland (3,100 \(km^2\)) (Table
\ref{tab-landcoverSurf}). The macrozones with major LULCC for 2001--2022
were ``Centro,'' ``Sur,'' and ``Austral'' with 36\%, 31\%, and 34\%,
respectively (Figure~\ref{fig-studyArea} and Table
\ref{tab-landcoverTrend}); of its surface that changes the type of land
cover. Figure~\ref{fig-LCprop} shows the summary of the proportion of
surface per land cover class and macrozone, derived from the persistance
mask over continental Chile.

@@

\hypertarget{land-cover-trend-and-drought-indices-1}{%
\subsubsection{Land cover trend and drought
indices}\label{land-cover-trend-and-drought-indices-1}}

@@

The ``Norte Chico'' shows an increase in barrend land of 111
\(km^2 year^{-1}\) and a reduction in the class savanna of 70
\(km^2 year^{-1}\). In the ``Centro'' and ``Sur,'' there are changes in
the Chilean matorral, with an important reduction in savanna (136 to 318
\(km^2 yr^{-1}\)), and an increase in shrubland and grassland. Showing a
change for more dense vegetation types. It appears to be a shift in the
area of cropland from the ``Centro'' to the ``Sur.'' Also, there is a
high increase in forest (397 \(km^2 yr^{-1}\)) in the ``Sur,'' replacing
the savanna lost.

Further, we want to address whether the trend in land cover change for
2001--2023 is associated with trends in drought indices of water demand
and supply and/or soil moisture for macrozone and land cover
macroclasses. From the three methods tested, Ridge, Lasso, and Random
Forest, neither gives significant results regarding whether the trend in
a drought index for any time scale explains the trend in land cover
change. Nevertheless, in ``Norte Chico'' and ``Centro,'' there is a
decrease in croplands and savanna and an increase in barren land, which
is associated with the variation in drought indices. Mainly for a
decrease in water supply (SPI and SSI) and an increase in water demand
(EDDI). However, due to the high variability from north to south in
Chile, the climatic conditions (arid, semi-arid, and humid), and the
land cover type, we believe that only in those zones could the LULCC be
driven to some degree by drought.

\hypertarget{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity-1}{%
\subsection{Trend of drought indices for water demand and supply, soil
moisture, and vegetation
productivity}\label{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity-1}}

In ``Norte Grande,'' vegetation productivity, as per the z-index,
exhibits a yearly increase of 0.02 in the five land cover macroclasses,
with respect to grassland and shrubland categories. There is a negative
trend in ``Norte Chico'' with -0.04 and ``Centro'' with -0.02 per
decade. In the ``Norte Chico,'' savanna (-0.05) has the lowest trend,
and the rest of the types are around -0.04. In ``Centro,'' shrubland
reaches -0.06, grassland -0.05, and croplands and savanna -0.01 per
decade. This could be associated either with a reduction in vegetation
surface, a decrease in biomass, or browning \citep{Miranda2023}.
Vegetation reached its lowest values since the year 2019, reaching an
extreme condition in early 2020 and 2022 in the ``Norte Chico'' and
Centro'' (Mega Drought). The ``Sur'' and ``Austral'' show a positive
trend of around 0.016 per decade (Figure~\ref{fig-zcNDVI_var}). Despite
the croplands suffering from drought just as badly as the native
vegetation in ``Norte Chico,'' the Chilean matorral \citep{Fuentes2021}
appears to be the region most affected by a negative trend in vegetation
productivity.

@@

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, breakable, interior hidden, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

\hypertarget{gridded-meteorological-and-vegetation-data}{%
\subsubsection{Gridded meteorological and vegetation
data}\label{gridded-meteorological-and-vegetation-data}}

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
The IGBP defines 17 classes; from these, we regrouped into ten
macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9
to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands,
13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
Thus, we have a land cover raster time series with the ten classes for
2001 and 2023. Prior to using this class, we compared it with the types
of land cover made by \citet{Zhao2016}, which is a more detailed land
cover map of Chile with a 30 m spatial resolution for 2012--2013. It
reached a global accuracy of \textasciitilde0.82. The procedure of
validation is demonstrated in Section S1 of the supplementary
material.To derive a proxy for vegetation productivity, we used the
MOD13A3 collection 6.1 product from MODIS \citep{Didan2015}. It provides
vegetation indices (NDVI and EVI) at 1km of spatial resolution and
monthly frequency. The MOD13A3 and MCD12Q1 were retrieved from the
online Data Pool, courtesy of the NASA EOSDIS Land Processes Distributed
Active Archive Center (LP DAAC), USGS Earth Resources Observation and
Science (EROS) Center, Sioux Falls, South Dakota,
https://lpdaa.usgs.gov/tools/data-pool/.

@@

For soil moisture, water supply, and water demand variables, we used
ERA5L \citep{MunozSabater2021}, a reanalysis dataset that provides the
evolution of the meteorological and soil moisture variables since 1950.
It has a spatial resolution of 0.1° (9 km), hourly frequency, and global
coverage. We selected the variables for total precipitation, maximum and
minimum temperature at 2 meters, and volumetric soil water layers
between 0 and 100cm of depth (layer 1 to layer 3).

\hypertarget{land-cover-change-and-trend}{%
\subsection{Land cover change and
trend}\label{land-cover-change-and-trend}}

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
6.1 from MODIS. \citet{Zambrano2018} and \citet{Fuentes2021} have
previously used this product for studies of drought and land cover. The
MCD12Q1 has a yearly frequency from 2001 to 2022. The IGBP defines 17
classes; from these, we regrouped into ten macroclasses, as follows:
classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as
grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as
snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land
cover raster time series with the ten macroclasses for 2001 and 2023. We
validate the land cover macroclasses regarding a highly detailed (30 m
of spatial resolution) land cover map made for Chile by \citet{Zhao2016}
for 2013-2014. Our results showed a global accuracy of
\textasciitilde0.82 and a F1 score of \textasciitilde0.66. Section S2 in
the Supplementary Material shows the procedure for validation.

Climate, vegetation development, seasonality, and changes in vegetation
type all have an impact on the time series of NDVI. In this study, we
want to examine the variation in vegetation productivity across various
land cover types and how water demand, water supply, and soil moisture
affect it. In order to avoid changes due to a change in the land cover
type that will wrongly impact NDVI, we developed a persistence mask for
land cover for 2001--2022. Thereby, we reduce an important source of
variation on a regional scale. Therefore, we generated a raster mask for
IGBP MODIS per pixel using macroclasses that remained unchanged for at
least 80\% of the years (2001--2022). This enabled us to identify
regions where the land cover macroclasses are persistent. We calculated
the surface occupied per land cover class into the five macrozones
(``Norte Grande'' to ``Austral'') per year for 2001--2023. After that,
we calculated the trend's change in surface per type. We used the Sen'
slope \citep{Sen1968} based on Mann-Kendall \citep{Kendall1975}.

\hypertarget{trend-of-drought-and-land-cover-change}{%
\subsection{Trend of drought and land cover
change}\label{trend-of-drought-and-land-cover-change}}

@@

In order to compute the drought indices that uses water demand, it is
necessary to first calculate the AED. To do this, we employed the
Hargreaves technique \citep{Hargreaves1994, Hargreaves1985}, by applying
the following equation:

@@

We chose the method of Hargreaves to estimate AED because of its
simplicity, which only requires temperatures and extrarrestrial
radiation. Also, it has been recommended over other methods (e.g.,
Penman-Monteith) when the access to climatic variables is limited
\citep{Vicente-Serrano2014}.

@@

To derive the drought indices of water supply and demand, soil moisture,
and vegetation we used the ERA5L dataset and the MODIS product, with a
monthly frequency for 1981-2023 and 2000-2023, respectively.

The dought indices correspond to a historical anomaly with regard to a
variable (e.g., meteorological, vegetation, or soil moisture). To
account for the anomaly, the common practice is to derive it following a
statistical parametric methodology in which it is assumed that the
statistical distribution of the data is known (\citet{Heim2002}). A
wrong decision is usually the highest source of uncertainty
(\citet{Laimighofer2022}). In the case of Chile, due to its high degree
of climatic variability, it is complex to choose a proper distribution
without previous research. Here, we follow a non-parametric methodology
for the calculation of the drought indices, in a similar manner as the
framework proposed by \citet{Farahmand2015};
\citet{Hobbins2016};\citet{McEvoy2016}.

For the purpose of monitoring water supply drought, we used the
well-known Standardized Precipitation Index (SPI), which the World
Meteorological Organization (WMO) recommended. The SPI solely relies on
precipitation data. Also, it has been used worldwide for the study of
drought, including in Chile (\citet{Garreaud2017};
\citet{Zambrano2017}). The primary cause of drought is precipitation
anomalies, and temperature usually makes it worse (Luo et al.~2017).
Nowadays, there is an increase in attention toward using water demand
separately to monitor droughts. (\citet{Vicente-Serrano2020};
\citet{Noguera2022}). Thus, to evaluate water demand, we chose the
Evaporative Demand Drought Index (EDDI), developed by Hobbins et
al.~(016) and McEvoy et al.~(2016), which is based on the AED. EDDI is
currently used for monitoring drought in the United States
(https://psl.noaa.gov/eddi/). In our case, we used only temperature for
AED, a difference from the original formulation of EDDI, which also
considered wind besides temperature. To consider the combined effect of
water supply and demand, we selected the SPEI, which corresponds to a
balance between precipitation and AED. \citet{Vicente-Serrano2010}
proposed the SPEI, and it has improved the SPI by incorporating
temperature for drought monitoring. For SPEI, an auxiliary variable D =
P-AED is calculated. Soil moisture is the main driver of vegetation
productivity, particularly in semi-arid regions (\citet{Li2022}). Hence,
for soil water drought, we used the SSI (Standardized Soil Moisture
Index) (Hao and AghaKouchak 2013; A. AghaKouchak 2014) which is a
multi-scale index similar to SPI, SPEI, and EDDI. In our case, for the
SSI, we used the average soil moisture from ERA5L at 1m depth. Finally,
for the proxy of productivity, we used the zcNDVI proposed by Zambrano
et al.~(2018) which will be derived from the NDVI retrieved from
MOD13A1.

To derive the drought indices, first we must calculate the sum of the
variables with regard to the time scale (s). In this case, for
generalization purposes, we will use \(V\), referring to variables
\(P\), \(AED\), \(D\), \(NDVI\), and \(SM\) (\#ref\{tab-desEOD\}). We
cumulated each \(V\) over the time series of \(n\) values (months), and
for the time scales \(s\):

@@

The \(A_{si}\) corresponds to a moving window (convolution) that sums
the variable for time scales \(s\) from the last month, month by month,
until the first month in which it could sum for \(s\) months. Once the
variable is cumulated over time for the scale \(s\). Thus, the
empirically derived probabilities are obtained through an inverse normal
approximation \citep{Abramowitz1968}. Then, we used the empirical Tukey
plotting position \citep{Wilks2011} over \(A_i\) to derive the
\(P(a_i)\) probabilities across a period of interest:

@@

The drought indices were calculated for time scales of 1, 3, 6, 12, 24,
and 36 months at a monthly frequency for 1981--2023 in order to be used
for short- to long-term evaluation of drought. In the case of the proxy
of vegetation productivity (zcNDVI) it was calculated for a time scale
of six months at monthly frequency for 2000--2023. For zcNDVI, we test
time scales of 1, 3, 6, and 12 months; we choose to use six months
because that shows a more robust representation of vegetation
productivity due to the seasonality of vegetation in Chile.

\hypertarget{trend-of-drought-indices}{%
\subsubsection{Trend of drought
indices}\label{trend-of-drought-indices}}

To estimate if there are significant positive or negative trends for the
drought indices, we used the non-parametric test of Mann-Kendall
\citep{Kendall1975}. To determine the magnitude of the trend, we used
Sen's slope \citep{Sen1968}. Some of the advantages of applying this
methodology are that the Sen's slope is not affected by outliers as
regular regression does, and it is a non-parametric method that is not
influenced by the distribution of the data. We applied Mann-Kendall to
see if the trend was significant and Sen's slope to estimate the
magnitude of the trend. We did this to the six time scales from 1981 to
2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI. Thus,
we have trends per index and time scale (24 in total). Then, we
extracted the trend aggregated by macrozone and per land cover persitent
macroclasses.

\hypertarget{drought-impacts-on-vegetation-productivity-within-land-cover}{%
\subsection{Drought impacts on vegetation productivity within land
cover}\label{drought-impacts-on-vegetation-productivity-within-land-cover}}

We analyzed the trend of vegetation productivity over the unchanged land
cover macroclasses. This way, we tried to reduce the noise in the
vegetation due to a change in land cover from year to year. To achieve
this, we will use the persistent mask of land cover macroclasses, which
are the types that remain more than 80\% of the time for 2001--2022. We
used this to evaluate the trend in zcNDVI per land cover class and
macrozone.

We examine the drought indices of water demand, water supply, soil
moisture, and their connection with vegetation productivity to
investigate two main questions: i) whether short-term or long-term time
scales have a greater impact on vegetation across Chile and its specific
regions; and ii) the spatial variation in the strength of the
correlation between the variables and time scales.~ Then, we will
summarize for each land cover class and macrozone. Thus, we will be able
to advance in understanding how climate is affecting vegetation,
considering the impact on the five macroclasses of vegetation: forest,
cropland, grassland, savanna, and shrubland.

We conducted an analysis on the linear correlation between the indices
SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36
months, and zcNDVI. The objective is to determine the impact of soil
moisture and water demand and supply on vegetation productivity. We used
a method similar to that used by \citet{Meroni2017} which compared the
SPI with the cumulative FAPAR (Fraction of Absorbed Photosynthetically
Active Radiation). A pixel-to-pixel linear correlation analysis was
performed for each index within the persistent mask of land cover
macroclasses. To begin, the Pearson coefficient of correlation is
computed for each of the six time scales. A significant time scale is
identified as the one that attains the highest correlation (p
\textless{} 0.05). Subsequently, the Pearson correlation coefficient
corresponding to the time scales at which the value peaked was
extracted. As a result, for each index, we generated two raster maps: 1)
containing the raster with values of the time scales that reached the
maximum correlation, and 2) having the value of the correlation
obtained.

@@

\hypertarget{land-cover-change-and-trend-1}{%
\section{Land cover change and
trend}\label{land-cover-change-and-trend-1}}

@@

For vegetation, we obtained and use hereafter five macroclasses of land
cover from IGBP MODIS: forest, shrubland, savanna, grassland, and
croplands. Figure~\ref{fig-studyArea}c shows the spatial distribution of
the macroclasses through Chile for the year 2022.
Figure~\ref{fig-studyArea}d shows the macroclasses of land cover
persistance (80\%) during 2021--2022, respectively
(\#ref\{tab-landcoverSurf\}). Within continental Chile, barren land is
the land cover class with the highest surface area (277,870 km2). The
largest type of vegetation, with 137,085 km2, is forest. Grassland
(74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland
(3,146 km2) are the other types (Table 2). The macrozones with major
changes for 2001--2022 were ``Centro,'' ``Sur,'' and ``Austral,'' with
36\%, 31\%, and 34\% of their surface changing the type of land cover,
respectively (Figure~\ref{fig-studyArea} and
\#ref\{tab-landcoverTrend\}). Figure~\ref{fig-LCprop} shows the summary
of the proportion of surface per land cover class and macrozone, derived
from the persistance mask over continental Chile.

@@

The ``Norte Chico'' shows an increase in barren land of 111
\(km^2 yr^{-1}\) and a reduction in the class savanna of 70
\(km^2 yr^{-1}\). In the ``Centro'' and ``Sur,'' there are changes with
an important reduction in savanna (136 to 318 \(km^2 yr^{-1}\) ), and an
increase in shrubland and grassland. Showing a change for more dense
vegetation types. It appears to be a shift in the area cultivated
(croplands) from the ``Centro'' to the ``Sur.'' Also, there is a high
increase in forest (397 \(km^2 yr^{-1}\) ) in the ``Sur,'' replacing the
savanna lost (\#ref\{tab-landcoverTrend\}).

\hypertarget{trend-on-drought-and-its-relation-to-land-cover-change}{%
\subsection{Trend on drought and its relation to land cover
change}\label{trend-on-drought-and-its-relation-to-land-cover-change}}

\hypertarget{trend-of-drought-indices-1}{%
\subsubsection{Trend of drought
indices}\label{trend-of-drought-indices-1}}

Figure~\ref{fig-trendDI} shows the spatial variation of the trend for
the drought indices from short- to long-term scales. SPI and SPEI have a
decreasing trend from ``Norte Chico'' to ``Sur.'' However, there is an
increasing trend in ``Austral.'' The degree of the trend is stronger at
higher time scales. The SSI indicates that in ``Norte Grande,'' there
are surfaces that have increased in the soutwest part and in the
northeast have decreased, and is shown for all time scales. Similar to
SPI and SPEI, SSI decreases at higher time scales. EDDI showed a
positive trend for the whole of continental Chile, with a higher trend
toward the north and a descending gradient toward the south. The degree
of trend increases at higher time scales.

@@

Figure~\ref{fig-corTimeScale} is a map that shows the highest
coefficient of determination (r-squared, or rsq) found in the regression
analysis between different drought indicators and plant productivity
over time. The spatial variation of time scales reached per index is
mostly for time scales above 12 months. In the case of SSI, the
predominant scales are 6 and 12 months. For all indices, to the north,
the time scales are higher and diminish toward the south until the south
part of ``Austral'' increases. In Figure~\ref{fig-corPerson}, the map of
Pearson correlation values is shown. The EDDI reached correlations above
0.5 between ``Norte Chico'' and ``Sur.'' The correlation changes from
negative to positive toward the Andes Mountains and to the sea, just as
in the northern part of ``Austral.'' The SPI and SPEI have similar
results, with the higher values in ``Norte Chico'' and ``Centro'' being
higher than 0.6. Following a similar spatial pattern as EDDI. The SSI
showed to be the index that has a major spatial extension with a higher
correlation. It has a similar correlation to SPI and SPEI for ``Norte
Chico'' and ``Sur,''~ but has improvements for ``Sur.''
